---
title: "Laboratory of Biostatistics Assessment"
author: "Group nº3 - Carolina Santos, Filipe Santos & Hugo Anjos"
date: "28 de Dezembro de 2019"
output:
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r dtStart, include=FALSE}
packages <- c("ggplot2", "here", "data.table", "unglue", "magrittr", "kableExtra", "knitr")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))
}
knitr::opts_chunk$set(out.width = "100%", echo = TRUE)

library(here)
library(data.table)
library(unglue)
library(ggplot2)
library(magrittr)
library(kableExtra)
```

# *Objectives*

The main objective is to *explore Liver and Intrahepatic Bile Ducts Cancer mortality geographical and time trends* for the following countries:

+ `Austria`

+ `Cyprus`

+ `Germany`

+ `Malta`

+ `Portugal`

+ `European Union`

To do so, we will answer the following questions, always *treating males and females separately*:

+ What are the time trends in the mortality rates between countries for all age groups combines? How do they compare with the European Union overall?

+ Compare the mortality rates by age groups between countries for 2016.

+ Which country changed (increase or decreased) the most in percentage terms between 2011 and 2016?

+ What are the five most common cancer in 2011 and in 2017?

# *Reading the Data*
```{r readind data}
data <- fread(here("CancerMortalityEU.csv"), header = TRUE, check.names = TRUE)
popdt <- fread("PopulationEU.csv", header = TRUE, check.names=TRUE)
```

# *Data Limitations*

By taking a quick look at the data sets we can observe that there is some missing data for some countries and for the year 2017, in specific, there is no data regarding the cancer for the countries of Portugal and Malta.

Also, there are some inherent limitations to the data in the sense that, the change registered in the mortality rate, from year to year, may be due to:

* Improvements to the detection methods for this kind of cancer

* An increase in the portion of the population being tested and diagnosed

Differences between countries might bue due to these limitations.

# *Data Tidying* 
## `Cancer Data`:

```{r cancerData, results= FALSE}
# Removing unnecessary columns
cancro <- data[CancerSite %in% "Liver and Intrahepatic Bile Ducts"]

cancro[, c("CancerSite", "Cause", "ageStart", "ageLabel"):=NULL]
setnames(cancro, "ageGroup", "Age")

# Melt of tha cancer data

cancer.new <- melt(cancro,
                   id.vars = c("Country","Age"),
                   variable.name = "all",
                   value.name = "Deaths")

cancer.new[, `:=` (Year=as.numeric(gsub("[^[:digit:].]",
                            "", all)))]

cancer.new[, Sex:=as.character(unglue_vec(all,"{x}_{y}",var="x"))]

cancer.new[, .N, by="Year"]
cancer.new[, .N, by="Sex"]
cancer.new[, all:=NULL]
setcolorder(cancer.new, c("Country","Age","Sex","Year","Deaths"))

#Removing Na
cancer.new <- na.omit(cancer.new)

# Criação do EU

cancer.EU <- cancer.new[, .(Country = "European Union", Deaths = sum(Deaths, na.rm = TRUE)), by = .(Year, Sex, Age)]
setcolorder(cancer.EU, c("Country","Age","Sex","Year","Deaths"))
cancer.liver <- rbind(cancer.new,cancer.EU)

# Criação do all

allAge <- cancer.liver[, .(Age = "All", Deaths = sum(Deaths, na.rm = TRUE)), by = .(Country, Year, Sex)]
setcolorder(allAge, c("Country", "Sex", "Age", "Year", "Deaths"))
cancer.liver <- rbind(cancer.liver,allAge)

# Selecting the important countries

cancer.liver <- cancer.liver[Country %in% c("Austria", "Cyprus", "Germany", "Malta", "Portugal", "European Union")
                             ][order(Year, Country, Sex)]
cancer.liver
```


## `Population Data`:

```{r dtPop, results= FALSE, warning=FALSE}
# Importing data
library(data.table)
popdt <- fread("PopulationEU.csv",
               header = TRUE, check.names=TRUE)

# Removing ageStart and ageLabel columns
popdt[, c("ageStart","ageLabel","X2000","X2001","X2002","X2003","X2004","X2005","X2006","X2007","X2008","X2009","X2010","X2018"):=NULL]

# Renaming ageGroup column to Age
setnames(popdt, c("ageGroup"),
                c("Age"))

# Morphing to long format
popT <- melt(popdt,
            id.vars = c("Country", "Sex", "Age"),
            measure = patterns("^X"),
            variable.name = "YearX",
            value.name = c("Population"),
            na.rm = TRUE)

# Year as numeric, removing YearX
popT[, `:=` (Year=as.numeric(gsub("[^[:digit:].]",
                            "", YearX)))][, YearX := NULL]
setcolorder(popT, c("Country", "Sex", "Age", "Year", "Population"))

# Removing rows for total population data 

#Seeing as we are treating females and males differently, the rows regarding total population info will not be used and should be removed.
popT <- popT[Sex != "Total"]

# Creating European Union as "Country"
#EU's population is the sum of the populations for each country, by year and age group

popEU <- popT[, .(Population = sum(Population, na.rm = TRUE)), by = .(Year, Sex, Age)][, Country := "European Union"]
setcolorder(popEU, c("Country", "Sex", "Age", "Year", "Population"))
pop <- rbind(popT, popEU)
pop <- pop[order(Year, Country)]

# Creating age group category "All", combining all ages
allAge <- pop[, .(Age = "All", Population = sum(Population, na.rm = TRUE)), by = .(Country, Year, Sex)]
setcolorder(allAge, c("Country", "Sex", "Age", "Year", "Population"))
allAge <- allAge[order(Year, Country)]

# Merging pop and allAge
pop <- rbind(pop, allAge)
pop <- pop[order(Year, Country, Sex)]

# Selecting only the countries in study
popF <- pop[Country %in% c("Austria", "Cyprus", "Germany", "Malta", "Portugal", "European Union")]
popF <- popF[order(Year, Country)]

popF
```

## `Merging the two data sets`:

```{r final data set, results=FALSE}
data.f <- cancer.liver[popF, on=c("Country","Age","Sex","Year")]
data.f[, mrate := round(100000*Deaths/Population,2)]
data.f[Age %in% "All"][Sex %in% "Females"][, sum(Deaths, na.rm = TRUE), by=Country]
data.f
```

## `Preparing the Data for the second question`:

```{r data.ex2, results=FALSE}
data.2016 <- data.f[Year %in% "2016"][Deaths != "NA"]
data.2016$Age <- as.factor(data.2016$Age)
levels(data.2016$Age)
data.2016$Age <- factor(data.2016$Age, levels(data.2016$Age)[c(1,10,2:9,11:19)])
```

## `Preparing the data for the third question`:

```{r dataEx3, results=FALSE}
# For the male population
data.diff.M <- data.f[Year != 2017][Year != 2011 & Sex %in% "Males" & Age %in% "All"]
data.diff.M[,per := ((data.diff.M$mrate - data.f[Year %in% 2011 & Sex %in% "Males" & Age %in% "All"]$mrate )/data.f[Year %in% 2011 & Sex %in% "Males" & Age %in% "All"]$mrate*100)]
diff.M <- data.diff.M[, DiffCountry := sprintf("%3.1f%%", per)][, .(Country, Year, DiffCountry)]
diff.Males <- dcast(diff.M, Year ~ Country, value.var = "DiffCountry" )

# For the female population

data.diff.F <- data.f[Year != 2017][Year != 2011 & Sex %in% "Females" & Age %in% "All"]
data.diff.F[,per := ((data.diff.F$mrate - data.f[Year %in% 2011 & Sex %in% "Females" & Age %in% "All"]$mrate )/data.f[Year %in% 2011 & Sex %in% "Females" & Age %in% "All"]$mrate*100)]
diff.F <- data.diff.F[, DiffCountry := sprintf("%3.1f%%", per)][, .(Country, Year, DiffCountry)]
diff.Females <- dcast(diff.F, Year ~ Country, value.var = "DiffCountry" )

```

## `Preparing the data for the fourth question`:

```{r allCancerSites, results=FALSE}
# Reading all cancer data
allCancer <- fread("CancerMortalityEU.csv",
               header = TRUE, check.names=TRUE)

# Removing superfluous columns
allCancer[, c("Cause", "ageStart", "ageLabel"):=NULL]
setnames(allCancer, "ageGroup", "Age")

# Converting table to long format
allCancer <- melt(allCancer,
                  id.vars = c("Country","Age","CancerSite"),
                  variable.name = "Sex+Year",
                  value.name = "Deaths")

# Separating Year and Sex data
allCancer[, `:=` (Year=as.numeric(gsub("[^[:digit:].]",
                            "", `Sex+Year`)))]
allCancer[, Sex:=as.character(unglue_vec(`Sex+Year`,"{x}_{y}",var="x"))]
allCancer[, "Sex+Year":=NULL]
setcolorder(allCancer, c("Country","Year","Sex","Age","CancerSite","Deaths"))

# Creating EU as a country
cancerEU <- allCancer[, .(Country = "European Union", Deaths = sum(Deaths)), by = .(Year, Sex, Age, CancerSite)]
setcolorder(cancerEU, c("Country","Year","Sex","Age","CancerSite","Deaths"))
allCancer <- rbind(allCancer, cancerEU)

# Creating All ages rows
allAge <- allCancer[, .(Age = "All", Deaths = sum(Deaths, na.rm = TRUE)), by = .(Country, Year, Sex, CancerSite)]
setcolorder(allAge, c("Country","Year","Sex","Age","CancerSite","Deaths"))
allCancer <- rbind(allCancer, allAge)

# Selecting the important countries
allCancer <- allCancer[Country %in% c("Austria", "Cyprus", "Germany", "Malta", "Portugal", "European Union")
                             ][order(Year, Country)]

# Removing "All Cancer" from CancerSite
allCancer <- allCancer[CancerSite != "All Cancers"]

allCancer
```


# *Questions*
## `First Question`

> What are the time trends in the mortality rates between countries for all age groups (“All”) combined?

>How do they compare with the European Union overall?

```{r Time trends for females, warning=FALSE, results=FALSE}
p <- data.f[Age %in% "All"][Sex %in% "Females"][Country != "Germany"] %>% 
ggplot(aes(Year,mrate, group=Country, colour = Country)) +
    geom_point(size=2) +
    geom_line() +
    ggtitle("Time trends of Liver Cancer on the Female population") +
    ylab("Mortality Rate") +
    scale_x_continuous(labels=data.f[,Year], breaks = data.f[,Year])+
    geom_point(data = data.f[Age %in% "All"][Sex %in% "Females"][Country %in% "Germany"], aes(Year, mrate, group= Country), colour = "yellow")+
    geom_line(data = data.f[Age %in% "All"][Sex %in% "Females"][Country %in% "Germany"], aes(Year, mrate, group= Country), colour = "yellow")

ggplot_build(p)$data

data.f[Age %in% "All"][Sex %in% "Females"] %>% 
ggplot(aes(Year,mrate, group=Country, colour = Country)) +
    geom_point(size=2) +
    geom_line() +
    ggtitle("Time trends of Liver Cancer on the Female population") +
    ylab("Mortality Rate") +
    scale_x_continuous(labels=data.f[,Year], breaks = data.f[,Year])+
    scale_colour_manual(labels = c("Austria", "Cyprus", "European Union", "Germany", "Malta", "Portugal"), values = c("#F8766D", "#B79F00", "#00BA38", "yellow", "#619CFF", "#F564E3"))
```
  
## *Interpretation*

  * There is an overall increasing trend for every country, as well as for the European Union.
  * Malta, Austria and Cyprus display the most irregular trends, probably explained by the lack of values for some of the ages in these countries.
    + `Austria` - with increases and decreases over the years, the mortality rate has been relatively stable, from 2011 to 2016.
    + `Malta` - an increase from 2011 (below 3) to 2016 (3,5), with a max in the year 2014. However, it remains one of the countries with the lowest mortality rate.
    + `Cyprus` - overall increase, with mortality peaks in 2013 and 2016.
  * Portugal, Germany and EU show an overall increase, with Portugal's mortality rate increasing the most among them (5.2 to 6).
  * These five countries show lower mortality rates than the European Union for every year, except Austria in 2012 and 2014.

```{r Time trends for males, warning=FALSE}
data.f[Age %in% "All"][Sex %in% "Males"] %>% 
ggplot(aes(Year,mrate, group=Country, colour = Country))+
    geom_point(size=2)+
    geom_line()+
    ggtitle("Time trends of Liver Cancer on the Male population")+
    ylab("Mortality Rate") +
    scale_x_continuous(labels=data.f[,Year], breaks = data.f[,Year])+
    scale_colour_manual(labels = c("Austria", "Cyprus", "European Union", "Germany", "Malta", "Portugal"), values = c("#F8766D", "#B79F00", "#00BA38", "yellow", "#619CFF", "#F564E3"))
```

## *Interpretation*

  * For males, like females, there is an increase in hepatic liver mortality rate over the years.
    + An exception is Germany, showing stable values for the years 2011 and 2017.
  * Malta displays the steepest increase in 2016 (from 7 to 13.8).
  *  Portugal displays the highest values (close to 17.5) while Cyprus has the lowest (even though the increase in mortality rate over the years is on par with Portugal).
  * The overall values for mortality are higher than those of females.
  * Portugal and Austria display higher values than the European Union, while the remaining countries stay below the average.


## `Second Question`

> Compare the mortality rates by age groups between countries for 2017.

```{r ageMR_Females}
data.2016[Sex %in% "Females"][Age != "All"] %>%
ggplot(aes(Age,mrate, group=Country, colour=Country))+
    geom_point()+
    geom_line()+
    theme(axis.text.x = element_text(angle=45, vjust = 0.5))+
    ggtitle("Mortlity rate per Age group in 2016 for the Female population")+
    ylab("Mortality Rate")+
    scale_colour_manual(labels = c("Austria", "Cyprus", "European Union", "Germany", "Malta", "Portugal"), values = c("#F8766D", "#B79F00", "#00BA38", "yellow", "#619CFF", "#F564E3"))
```

## *Interpretation*

  * In every country (except in Malta which is a lot harder to describe) we see a general increase on the mortality rate as the age class increases (starting in the 35 to 39 age class)
  * Contrary to the other countries (which always increase), Malta's mortality rate oscilates a lot between the classes 55-60 and 85, where it ends with a 0 mortality rate
  * The first Age classes (until the 35-39 age class) show zero to low mortality rate for this type of cancer on the Female population
  * The trend observed here is that the older population (from the age of 50 foward) is more susceptible to die due to this kind of cance. Germany, Austria, Portugal, Malta and Cyprus have a lower mortality rate when compared to the mortality rate of the European Union
    + Malta only has a higher mortatlity rate than the European Union in the classes of 60-64 and 80-84
    + Cyprus only has a higher mortality rate than the European Union in the classes of 50-54, 65-69, 80-84 and 85


```{r ageMR_Males}
data.2016[Sex %in% "Males"][Age != "All"] %>%
ggplot(aes(Age,mrate, group=Country, colour=Country))+
    geom_point()+
    geom_line()+
    theme(axis.text.x = element_text(angle=45, vjust = 0.5))+
    ggtitle("Mortlity rate per Age group in 2016 for the Male population")+
    ylab("Mortality Rate") +
    scale_x_discrete(labels=data.f[,Age], breaks = data.f[,Age])+
    scale_colour_manual(labels = c("Austria", "Cyprus", "European Union", "Germany", "Malta", "Portugal"), values = c("#F8766D", "#B79F00", "#00BA38", "yellow", "#619CFF", "#F564E3"))
```

## *Interpretation*

  * Again, in every country we see a general increase on the mortality rate as the age class increases (starting in the 40 to 44 age class)
  * Contrary to the general trend, we see a decrease in Malta between the age classes of 50-54 and 65-69m as well as with Cyprus between 55-59 and 60-64, 60-64 and 70-74 (which stabilizes) and 80-84 and 85 where it drop to 21.72
  * As well as with the female population, the first Age classes (until the class 40-44) show zero to very few mortality rate for this type of cancer on the Female population
  * Again, there is a trend for older people (from the age of 50 forward) to be more susceptible to die due to this kind of cancer. Germany, Austria, Portugal and Cyprus have a lower mortality rate than the European Union, while Malta has a higher mortality rate
    + However, Portugal has a higher mortality rate, when compared to the European Union, between the age classes of 50-54 to 65-69
    + Austria also shows a higher mortality rate than the European Union between the age classes of 60-64 to 75-79 
    + On the other hand, Malta only shows a mortality rate lower than the European Union in the age classes of 60-64 and 65-69
  * In general, the mortality rate for this type of cancer is higher in the Male population than on the Female population, for the countries chosen for analysis

## `Third Question`

> Which country changed (increase or decreased) the most in percentage terms between 2011 and 2016?

> Plot and tabulate the percentage change overtime by country.

```{r female population}

# For the female population
data.diff.F %>%
  ggplot(aes(Year, per, group=Country, colour =Country))+
    geom_point()+
    geom_line()+
    ggtitle("Mortality rate when compared with 2011 for the Female population")+
    ylab("Difference in %")+
    scale_colour_manual(labels = c("Austria", "Cyprus", "European Union", "Germany", "Malta", "Portugal"), values = c("#F8766D", "#B79F00", "#00BA38", "yellow", "#619CFF", "#F564E3"))

# For the Female population
diff.Females %>%
  kable(align="ccccc") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = T)
```

### *Interpretation*

The country that changed the most in mortality rate for the Female population, from 2011 to 2016, was Cyprus with an increase of 29.9%


```{r Male population}
# For the male population

diff.Males %>%
  kable(align="ccccc") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = T)

# For the male population
data.diff.M %>%
  ggplot(aes(Year, per, group=Country, colour =Country))+
    geom_point()+
    geom_line()+
    ylab("Difference in %")+
    scale_colour_manual(labels = c("Austria", "Cyprus", "European Union", "Germany", "Malta", "Portugal"), values = c("#F8766D", "#B79F00", "#00BA38", "yellow", "#619CFF", "#F564E3"))

```

### *Interpretation*

The country that changed the most in mortality rate for the Male population, from 2011 to 2016, was Malta with an increase of 71.5% 

## `Fourth Question`

> What are the five (5) most common cancers in 2011? in 2016 for your countries?

```{r cancersF11}
#Selecting cancer data for 2011
cancersF <- allCancer[Sex=="Females"][Year=="2011"][Age=="All"][order(CancerSite)]
cancersF <- cancersF[, .(Deaths = sum(Deaths, na.rm = TRUE)), by = .(CancerSite)]
cancersF[order(-Deaths), head(.SD,5)]%>%
  kable(align = "cc")%>% 
  kable_styling(bootstrap_options = "striped")
```

```{r cancersF16}
#Selecting cancer data for 2016
cancersF <- allCancer[Sex=="Females"][Year=="2016"][Age=="All"][order(CancerSite)]
cancersF <- cancersF[, .(Deaths = sum(Deaths, na.rm = TRUE)), by = .(CancerSite)]
cancersF[order(-Deaths), head(.SD,5)]%>%
  kable(align = "cc")%>% 
  kable_styling(bootstrap_options = "striped")
```

## *Interpretation*

For the countries chosen (Austria, Cyprus, Germany, Malta, Portugal and the European Union), the **5 most common cancer types were the same in 2011 and 2016: Breast, Trachea/Bronchus/Lung, Colon, Pancreas and Ovary**. There is a slight increase in the number of deaths for Breast, Trachea/Bronchus/Lung and Pancreas cancers (Pancreas showing the largest increase), while the mortality for Colon and Ovary shows a slight decrease.


```{r cancersM11}
#Selecting cancer data for 2011
cancersM <- allCancer[Sex=="Males"][Year=="2011"][Age=="All"][order(CancerSite)]
cancersM <- cancersM[, .(Deaths = sum(Deaths, na.rm = TRUE)), by = .(CancerSite)]
cancersM[order(-Deaths), head(.SD,5)]%>%
  kable(align = "cc")%>% 
  kable_styling(bootstrap_options = "striped")
```

```{r cancersM16}
#Selecting cancer data for 2016
cancersM <- allCancer[Sex=="Males"][Year=="2016"][Age=="All"][order(CancerSite)]
cancersM <- cancersM[, .(Deaths = sum(Deaths, na.rm = TRUE)), by = .(CancerSite)]
cancersM[order(-Deaths), head(.SD,5)] %>%
  kable(align = "cc")%>% 
  kable_styling(bootstrap_options = "striped")
```

## *Interpretation*

For males, the **most common cancers remained mostly the same** between the years of 2011 and 2016, the only difference being the **increase in deaths caused by Liver and Intrahepatic Bile Ducts cancer**, overtaking Stomach cancer as the fifth most deadly type of cancer. The **most common cancer** cause of death for both years was **Trachea/Bronchus/Lungs** by a large margin. However, it is the only cancer type that shows a decrease in numbers from 2011 to 2016. The remaining cancer types show an increase in the number of deaths, with Pancreas cancer displaying the largest.

# *Conclusions*:

 * This data shows an **overall slight increase in the mortality rate for Liver and Intrahepatic Bile Ducts cancer** (from 2011 to 2016)
  
 * **Lack of data for 2017** makes this analysis more difficult
  
 * The data also shows that this cancer is more aggressive to older people, showing **higher mortality rates for higher age groups**.
  
 * **Most common cancers are** (almost) **the same** from 2011 to 2016. There is a slight overall trend for the number of deaths to increase
  
 * Given the data limitations, the **true mortality rate may be higher**

```{r sessionInfo, echo=TRUE}
sessionInfo()
```

